\bf Looking at Non-Linear Dimension Reductions as Models in the Data Space

### 2024-07-21 paper.qmd
Removed long edges after inspecting the model in p-D
Update choice of bins section

The performance and robustness of our model depend on three key parameters: (i) the total number of bins ($b$), (ii) a benchmark value used to remove low-density hexagons, and (iii) a benchmark value used to remove long edges. However, there is no analytical formula to calculate an appropriate value for these parameters. The selection of these parameter values depends on the model performance computed by MSE (see @sec-summary).

#### Choice of bins

The number of hexagon bins in the hexagon grid has a considerable impact on the construction of the \gD{} model, serving as the initial step. The chosen total number of bins must effectively capture the structure of the NLDR data. If the number of bins is too low, the model may not be able to capture the structure of the NLDR data effectively (see @fig-bins-scurve (a)), while if there are too many bins, it may result in over-fitting the individual points of the NLDR data (see @fig-bins-scurve (c)). Therefore, it is important to determine an effective number of bins to construct a reasonable model (see @fig-bins-scurve (b)).

@@

```{r}
#| echo: false
#| label: fig-bins-scurve
#| fig-pos: H
#| fig-cap: "Hexbin density plots of UMAP layout of the S-curve data, using three different bin inputs: (a) $b = 91 \\text{ } (7, \\text{ }13)  \\text{ }(q = 0.12)$, (b) $b = 220 \\text{ } (11, \\text{ }20) \\text{ }(q = 0.07)$, and (c) $b = 527 \\text{ } (17, \\text{ }31) \\text{ }(q = 0.05)$. Color indicates standardized counts, dark indicating high count and light indicates low count. At the smallest bin size the data segregates into two separate groups, suggesting this is too many bins. Using the MSE of the model fit in $p-D$ helps decide on a useful choice of number of bins."
#| fig-width: 6
#| fig-height: 3

@@

To determine the effective $b$, candidate values are selected based on the range between the minimum and approximate maximum $b_1$, because $b_2$ is computed from $b_1$. The minimum $b_1$ is set to $2$, while the maximum number is estimated by taking the square root of $\frac{n}{2}$. The analysis evaluates the MSE across varying $b$ within this range, covering the minimum to maximum values along both axes (see @fig-mse-scurve-b).

@@

<!--To generate errors for different total number of bins-->
```{r}
## To initialize number of bins along the x-axis
bin1_vec_scurve <- 2:19 #sqrt(NROW(training_data_scurve)/2)

@@

for (xbins in bin1_vec_scurve) {

  bin2 <- calc_bins_y(bin1 = xbins, r2 = r2, q = 0.07)$bin2

  scurve_model <- fit_highd_model(
    training_data = training_data_scurve,
    emb_df = umap_scurve_scaled,
    bin1 = xbins,
    r2 = r2,
    q = 0.07, 
    is_bin_centroid = TRUE,
    is_rm_lwd_hex = FALSE,
    col_start_highd = "x"
  )

@@

  ## Compute error
  error_df <- glance(
    df_bin_centroids = df_bin_centroids_scurve,
    df_bin = df_bin_scurve,
    training_data = training_data_scurve,
    newdata = NULL,
    type_NLDR = "UMAP",
    col_start = "x") |>
    mutate(bin1 = xbins,
           bin2 = bin2,
           b = bin1 * bin2,
           b_non_empty = NROW(df_bin_centroids_scurve))

@@

mse_scurve_b <- ggplot(error_scurve, 
                     aes(x = b, 
                         y = log(MSE))) + 
  geom_point(size = 1) +
  geom_line() + 
  geom_vline(xintercept = 220, linetype="solid", 
             color = "black", linewidth=0.8, alpha = 0.5) +
  geom_vline(xintercept = 112, linetype=2, 
             color = "black", linewidth=0.8, alpha = 0.5) +
  geom_vline(xintercept = 364, linetype=2, 
             color = "black", linewidth=0.8, alpha = 0.5) +
  ylab("log(MSE)") +
  xlab("b") +
  ggtitle("(a)") +
  theme(aspect.ratio = 0.75,
        axis.text.x = element_text(size = 5),
        axis.text.y = element_text(size = 5),
        axis.title.x = element_text(size = 5),
        axis.title.y = element_text(size = 5, 
            angle=90),
        title = element_text(size = 72))
```

```{r}
#| fig-cap: "MSE from UMAP applied to S-curve dataset with diffferent $b$ choices. What is the effective $b$ to create the model? The residual plot have a steep slope at the beginning, indicating that a smaller $b$ causes a larger amount of MSE. Then, the slope gradually declines or level off, indicating that a higher $b$ generates a smaller MSE. Using the elbow method, it was observed that when the $b$ is set to $190$, the lowest MSE occurred."
#| label: fig-mse-scurve-b
#| fig-pos: H
#| out-height: 50%
#| eval: false

mse_scurve_b
```

@@

```{r}
#| echo: false
#| fig-cap: "Various plots to help assess best number of bins, low density bin and large edge removal. Both (a) and (c) show MSE, against number of bins and standardised count. A good benchmark value for these parameters is when the MSE drops and then flattens out. Plot (b) shows the distribution of stadardised counts of hexagons. Plot (c) shows the distribution of $2\\text{-}D$ Euclidean distances between bin centroids, with a good benchmark value for removing large edges would being the distance that shows the first large increase."

#| label: fig-dist-scurve
#| out-width: 100%
#| fig-width: 12
#| fig-height: 10
#| fig-pos: H

@@

>>>>

The performance and robustness of our model depend on four key parameters: (i) bin start position $(s_1, \ s_2)$, (ii) the total number of bins ($b$), (iii) a benchmark value used to remove low-density hexagons, and (iv) a benchmark value used to remove long edges. However, there is no analytical formula to calculate appropriate values for these parameters. The selection of these parameter values depends on the model performance computed by MSE (@sec-summary), except for the threshold used to remove long edges.

#### Bin start position and number of bins

The starting position of the hexagonal grid is important because different starting points result in different distributions of data across bins, even with the same total number of bins. This variation affects the model due to the differing number of non-empty bins. Therefore, it is necessary to evaluate various starting points with different total numbers of bins to determine which configurations are more effective at capturing the structure and fitting the model.

@@

```{r}
#| echo: false
#| label: fig-bins-scurve
#| fig-pos: H
#| fig-cap: "Hexbin density plots of UMAP layout of the S-curve data, using three different bin inputs: (a) $b = 91 \\text{ } (7, \\text{ }13)  \\text{ }(q = 0.12)$, (b) $b = 220 \\text{ } (11, \\text{ }20) \\text{ }(q = 0.07)$, and (c) $b = 527 \\text{ } (17, \\text{ }31) \\text{ }(q = 0.05)$. Color indicates standardized counts, dark indicating high count and light indicates low count. At the smallest bin size the data segregates into two separate groups, suggesting this is too many bins. Using the MSE of the model fit in $p\\text{-}D$ helps decide on a useful choice of number of bins."
#| fig-width: 6
#| fig-height: 3

@@

To determine the effective $b$, candidate values are selected based on the range between the minimum and approximate maximum $b_1$, because $b_2$ is computed from $b_1$. The minimum $b_1$ is set to $2$, while the maximum number is estimated by taking the square root of $\frac{n}{2}$. By evaluating MSE across varying $b$ within this range for different $q$, helps to determine an appropriate values for $b$ and $q$ (@fig-param-scurve (a)).

@@

<!--To generate errors for different total number of bins-->
```{r}
## To initialize number of bins along the x-axis
bin1_vec_scurve <- 2:19 #sqrt(NROW(training_data_scurve)/2)
buffer_vec <- seq(0.05, 0.1, by = 0.01)

@@

for (xbins in bin1_vec_scurve) {
  for (buff in buffer_vec) {

    bin2 <- calc_bins_y(bin1 = xbins, r2 = r2, q = buff)$bin2

    scurve_model <- fit_highd_model(
      training_data = training_data_scurve,
      emb_df = umap_scurve_scaled,
      bin1 = xbins,
      r2 = r2,
      q = buff, 
      is_bin_centroid = TRUE,
      is_rm_lwd_hex = FALSE,
      col_start_highd = "x"
    )

@@

    ## Compute error
    error_df <- glance(
      df_bin_centroids = df_bin_centroids_scurve,
      df_bin = df_bin_scurve,
      training_data = training_data_scurve,
      newdata = NULL,
      type_NLDR = "UMAP",
      col_start = "x") |>
      mutate(bin1 = xbins,
             bin2 = bin2,
             b = bin1 * bin2,
             b_non_empty = NROW(df_bin_centroids_scurve),
             buff = buff)

@@

mse_scurve_b <- ggplot(error_scurve, 
                     aes(x = b, 
                         y = log(MSE),
                         color = buff,
                         group = buff)) + 
  geom_point(size = 0.05) +
  geom_line(linewidth = 0.05) + 
  geom_vline(xintercept = 220, linetype="solid", 
             color = "black", linewidth=0.1) +
  geom_vline(xintercept = 112, linetype=2, 
             color = "black", linewidth=0.1) +
  geom_vline(xintercept = 364, linetype=2, 
             color = "black", linewidth=0.1) +
  scale_colour_continuous_sequential(palette = "Magenta") +
  labs(x = expression(b), y = "log(MSE)", color = expression(q)) +
  ggtitle("(a)") +
  theme(aspect.ratio = 0.75,
        axis.text.x = element_text(size = 5),
        axis.text.y = element_text(size = 5),
        axis.title.x = element_text(size = 5),
        axis.title.y = element_text(size = 5, 
            angle=90),
        title = element_text(size = 72),
        legend.title = element_text(size = 5),
        legend.text = element_text(size = 3))
```

@@

```{r}
#| echo: false
#| fig-cap: "Various plots to help assess best bin start position, number of bins, low density bin and large edge removal. Both (a) and (c) show MSE, against number of bins and standardised count, while (a) also shows the change in MSE for different $q$. A good benchmark value for these parameters is when the MSE drops and then flattens out. Plot (b) shows the distribution of stadardised counts of hexagons. Plot (d) shows the distribution of $2\\text{-}D$ Euclidean distances between bin centroids, with a good benchmark value for removing large edges would being the distance that shows the first large increase."
#| label: fig-param-scurve
#| out-width: 100%
#| fig-width: 12
#| fig-height: 10
#| fig-pos: H

@@

## Remove an edge that exist as long edge in p-D
# tr_from_to_df_pbmc |> mutate(ID = row_number()) |> filter(from == 37) |> filter(to == 43) 
tr_from_to_df_pbmc <- tr_from_to_df_pbmc |>
  filter(row_number() != 360)

### 2024-07-21 paper.tex
Update choice of bins section

The performance and robustness of our model depend on three key
parameters: (i) the total number of bins (\(b\)), (ii) a benchmark value
used to remove low-density hexagons, and (iii) a benchmark value used to
remove long edges. However, there is no analytical formula to calculate
an appropriate value for these parameters. The selection of these
parameter values depends on the model performance computed by MSE (see
Section~\ref{sec-summary}).

\subsubsection{Choice of bins}\label{choice-of-bins}

The number of hexagon bins in the hexagon grid has a considerable impact
on the construction of the \gD{} model, serving as the initial step. The
chosen total number of bins must effectively capture the structure of
the NLDR data. If the number of bins is too low, the model may not be
able to capture the structure of the NLDR data effectively (see
Figure~\ref{fig-bins-scurve} (a)), while if there are too many bins, it
may result in over-fitting the individual points of the NLDR data (see
Figure~\ref{fig-bins-scurve} (c)). Therefore, it is important to
determine an effective number of bins to construct a reasonable model
(see Figure~\ref{fig-bins-scurve} (b)).

@@

\caption{\label{fig-bins-scurve}Hexbin density plots of UMAP layout of
the S-curve data, using three different bin inputs: (a)
\(b = 91 \text{ } (7, \text{ }13)  \text{ }(q = 0.12)\), (b)
\(b = 220 \text{ } (11, \text{ }20) \text{ }(q = 0.07)\), and (c)
\(b = 527 \text{ } (17, \text{ }31) \text{ }(q = 0.05)\). Color
indicates standardized counts, dark indicating high count and light
indicates low count. At the smallest bin size the data segregates into
two separate groups, suggesting this is too many bins. Using the MSE of
the model fit in \(p-D\) helps decide on a useful choice of number of
bins.}

@@

To determine the effective \(b\), candidate values are selected based on
the range between the minimum and approximate maximum \(b_1\), because
\(b_2\) is computed from \(b_1\). The minimum \(b_1\) is set to \(2\),
while the maximum number is estimated by taking the square root of
\(\frac{n}{2}\). The analysis evaluates the MSE across varying \(b\)
within this range, covering the minimum to maximum values along both
axes (see \textbf{?@fig-mse-scurve-b}).

@@

{\centering \includegraphics[width=1\textwidth,height=\textheight]{paper_files/figure-pdf/unnamed-chunk-31-1.pdf}

@@

\caption{Various plots to help assess best number of bins, low density
bin and large edge removal. Both (a) and (c) show MSE, against number of
bins and standardised count. A good benchmark value for these parameters
is when the MSE drops and then flattens out. Plot (b) shows the
distribution of stadardised counts of hexagons. Plot (c) shows the
distribution of \(2\text{-}D\) Euclidean distances between bin
centroids, with a good benchmark value for removing large edges would
being the distance that shows the first large increase.}

>>>>

The performance and robustness of our model depend on four key
parameters: (i) bin start position \((s_1, \ s_2)\), (ii) the total
number of bins (\(b\)), (iii) a benchmark value used to remove
low-density hexagons, and (iv) a benchmark value used to remove long
edges. However, there is no analytical formula to calculate appropriate
values for these parameters. The selection of these parameter values
depends on the model performance computed by MSE
(Section~\ref{sec-summary}), except for the threshold used to remove
long edges.

\subsubsection{Bin start position and number of
bins}\label{bin-start-position-and-number-of-bins}

The starting position of the hexagonal grid is important because
different starting points result in different distributions of data
across bins, even with the same total number of bins. This variation
affects the model due to the differing number of non-empty bins.
Therefore, it is necessary to evaluate various starting points with
different total numbers of bins to determine which configurations are
more effective at capturing the structure and fitting the model.

@@

\caption{\label{fig-bins-scurve}Hexbin density plots of UMAP layout of
the S-curve data, using three different bin inputs: (a)
\(b = 91 \text{ } (7, \text{ }13)  \text{ }(q = 0.12)\), (b)
\(b = 220 \text{ } (11, \text{ }20) \text{ }(q = 0.07)\), and (c)
\(b = 527 \text{ } (17, \text{ }31) \text{ }(q = 0.05)\). Color
indicates standardized counts, dark indicating high count and light
indicates low count. At the smallest bin size the data segregates into
two separate groups, suggesting this is too many bins. Using the MSE of
the model fit in \(p\text{-}D\) helps decide on a useful choice of
number of bins.}

@@

To determine the effective \(b\), candidate values are selected based on
the range between the minimum and approximate maximum \(b_1\), because
\(b_2\) is computed from \(b_1\). The minimum \(b_1\) is set to \(2\),
while the maximum number is estimated by taking the square root of
\(\frac{n}{2}\). By evaluating MSE across varying \(b\) within this
range for different \(q\), helps to determine an appropriate values for
\(b\) and \(q\) (Figure~\ref{fig-param-scurve} (a)).

@@

\includegraphics[width=1\textwidth,height=\textheight]{paper_files/figure-pdf/fig-param-scurve-1.pdf}

@@

\caption{\label{fig-param-scurve}Various plots to help assess best bin
start position, number of bins, low density bin and large edge removal.
Both (a) and (c) show MSE, against number of bins and standardised
count, while (a) also shows the change in MSE for different \(q\). A
good benchmark value for these parameters is when the MSE drops and then
flattens out. Plot (b) shows the distribution of stadardised counts of
hexagons. Plot (d) shows the distribution of \(2\text{-}D\) Euclidean
distances between bin centroids, with a good benchmark value for
removing large edges would being the distance that shows the first large
increase.}
